Ignition of Fuel Beds by Firebrands

### 2022-05-24 introduction.tex
updated some formatting

    This work aims to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it. Four specific objectives were addressed to act as a framework for achieving the overall goal based on the current knowledge of fuel bed ignition. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of sizes of the recipient fuels on ignition behavior
            \item Ascertain ignition dependence on heating location(s), mode, and rate
            \item Identify and quantify the influence of wind on ignition propensity.
            \item Identify the influence of fuel bed chemical composition on flaming ignition.
        \end{enumerate}
    The structure of this dissertation is as follows. First the current state of knowledge of fuel bed ignition modelling is summarized as it applies to this work (Chapter~\ref{part:literature}). Literature review specific to each objective is contained in the corresponding manuscripts. The results of this effort are then presented in manuscript form, followed by the conclusions, suggestions for future work, and appendices. The first manuscript addresses sensitivities of ignition to the particle size of the fuel bed materials (Chapter~\ref{part:manuscript1}). The second manuscript explores the influence of wind and fluid phenomena around the embers in the presence of a heat source unaffected by wind (Chapter~\ref{part:manuscript2}). The third manuscript identifies sensitivities to ignition related to chemical composition and thermal properties in both quiescent conditions and with wind (Chapter~\ref{part:manuscript3}). The fourth manuscript identifies changes in ignition propensity when multiple embers are located in close proximity on the fuel bed (Chapter~\ref{part:manuscript1}). The final chapter summarizes the conclusions of each manuscript, discusses the implications of the conclusions with respect to predictive using an example ignition model, and suggests avenues for future work (Chapter~\ref{part:conclusion}).

    It is anticipated that this work will provide insights and a basis for the creation of a simplified model that can accurately predict ignition. For example, in a recent conversation with a developer of the Fire Dynamics Simulator (FDS) the need was expressed for an accurate and low computational cost methodology or model to predict ignition fro better predictions of fire spread. Due to the large scales in fire simulations (~10\si{\meter} grid size) a detailed model of ignition (<1\si{\milli\meter} grid size) is not feasible. Similarly, risk management personnel or homeowners are not likely to have the capability to create detailed models and predictions of risk around structures and homes they are tasked with protecting. The goal and perceived impact of this work is to create or enable the creation of models that can be effectively implemented by the people interacting with fires and their risks in the field.

>>>>

    This work aims to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it. Four specific objectives were addressed to act as a framework for achieving the overall goal based on the current knowledge of fuel bed ignition. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of sizes of the recipient fuels on ignition behavior
            \item Ascertain ignition dependence on heating location(s), mode, and rate
            \item Identify and quantify the influence of wind on ignition propensity.
            \item Identify the influence of fuel bed chemical composition on flaming ignition.
        \end{enumerate}
    The structure of this dissertation is as follows. First, the current state of knowledge of fuel bed ignition modeling is summarized as it applies to this work (Chapter~\ref{part:literature}). A literature review specific to each objective is contained in the corresponding manuscripts. The results of this effort are then presented in manuscript form, followed by the conclusions, suggestions for future work, and appendices. The first manuscript addresses sensitivities of ignition to the particle size of the fuel bed materials (Chapter~\ref{part:manuscript1}). The second manuscript explores the influence of wind and fluid phenomena around the embers in the presence of a heat source unaffected by wind (Chapter~\ref{part:manuscript2}). The third manuscript identifies sensitivities to ignition related to chemical composition and thermal properties in both quiescent conditions and with wind (Chapter~\ref{part:manuscript3}). The fourth manuscript identifies changes in ignition propensity when multiple embers are located close to the fuel bed (Chapter~\ref{part:manuscript1}). The final chapter summarizes the conclusions of each manuscript, discusses the implications of the conclusions with respect to predictive using an example ignition model, and suggests avenues for future work (Chapter~\ref{part:conclusion}).

    It is anticipated that this work will provide insights and a basis for the creation of a simplified model that can accurately predict ignition. For example, in a recent conversation with a developer of the Fire Dynamics Simulator (FDS) the need was expressed for an accurate and low computational cost methodology or model to predict ignition for better fire spread predictions. Due to the large scales in fire simulations (~10\si{\meter} grid size) a detailed model of ignition (<1\si{\milli\meter} grid size) is not feasible. Similarly, risk management personnel or homeowners are not likely to have the capability to create detailed models and predictions of risk around structures and homes they are tasked with protecting. The goal and perceived impact of this work is to create or enable the creation of models that can be effectively implemented by the people interacting with fires and their risks in the field.

### 2022-05-24 litreview.tex
updated some formatting

\chapter{Background}
\label{part:literature}
    This chapter provides a discussion of the background and motivation of this work in further depth than the manuscripts presented later in this dissertation.  Here, a general overview of ignitions as a threat to structures and fire spread as well as the current state of knowledge and predictive capabilities is communicated. Each manuscript includes a relevant and more detailed literature review pertaining to the specific objective of the chapter.

\section{Widlfires and the WUI}
    Wildfires have been an integral part of many ecosystems for millennia and are often beneficial or required for maintaining a healthy ecosystem. As beneficial as they may be to the natural environment wildfires pose significant risk to the built environment. As communities around the world have expanded communities outward into environments that support wildfires (e.g., forests and grasslands) the risk of destruction of structures by wildfires has increased~\cite{Hammer2009DemographicManagement}. In the United States, 91\% of the Wildland Urban Interface (WUI) \nomenclature{WUI}{Wildland Urban Interface} growth of 189,000\si{\kilo\meter\squared} (larger than the state of Washington) development between 1990 and 2010 is occupied by homes~\cite{Radeloff2017} and has continued to grow on recent years. The increased development in theses areas means that an excess of 40 million homes are located in areas that may be threatened by a wildfire.

    In addition to the increase of homes in the WUI the intensity and severity of fires has increased. California is an example of the increased fire severity where 13 of the 20 most destructive fires have occurred since 2010 and 7 of the 20 most destructive occurred in 2020 and 2021~\cite{CALFIRE2018}. Increasingly severe fires are not just limited to California. Texas and Tennessee each had one of the top ten most severe fires in the United States between 2005 and 2020. Colorado, Oregon, Washington, Oklahoma, and Florida have also seen highly destructive fires in recent years~\cite{Barrett2020}. The United States is not unique in the presence of wilfdires that destroy homes. Australia's Black Summer in 2019-2020 was one of the most severe fire seasons in the countries' history~\cite{Levin2021Unveiling2019/2020} where over 3000 homes and greater than 46 million acres burned~\cite{Filkov2020ImpactTrends}, which is approximately equivalent to 75\% of Oregon's land area. The Mediterranean countries of Portugal, Spain, France, Italy, and Greece are also either experiencing greater losses or are a risk for greater losses due to wildfires in recent years. Similar trends are also present in many Asian and South American countries~\cite{Manzello2018}. Wildfires that threaten lives, homes, and structures are truly a global phenomena and it is essential to understand how homes are ignited and destroyed in wildfires to better prevent losses from occurring.

\section{The Problem of Ignition}
    Ignition of a home in a WUI fire is attributed to three primary mechanisms; direct flame contact, ignition by radiative heating, and ignition by firebrands. Of these methods ignition by firebrands is considered to be a primary threat to homes~\cite{Suzuki2021, Mell2010, Manzello2018FORUMResearch}. Case studies of WUI fires have consistently attributed structural losses to firebrand ignition~\cite{Westhaver2017WhyDisaster, Roberts2021} with one study attributing at least 2/3 of home losses to firebrands~\cite{Mell2011}. With the majority of ignitions occurring due to firebrand ignition it is imperative to understand the mechanism of firebrand ignition to reduce home losses. An analysis of structures burned in California wildfires from 2013-2018 found that many of the homes lost would have been considered "fire-safe" by current standards but were still ignited by embers~\cite{Syphard2019}. Codes, standards, and best practices built upon a better understanding of ignition by firebrands would help provide better resilience to firebrand attacks~\cite{Manzello2020} reducing the number of homes lost in wildfires.  Examples of these changes may include adjustments in building and landscaping materials favoring the selection of harder to ignite species or housing geometry constraints to reduce collection of flammable materials or embers. Without a comprehensive knowledge of the ignition process, creating the codes and standards to reduce the risk of home loss is challenging at best and may even be counterproductive.

    The general mechanism for firebrand ignition is considered to be a three step process which starts within the fire itself, often at a significant distance from the ignition point. First, a firebrand is generated as burning material breaks off or is lofted into the air. Second, the firebrand is then lofted through the air and transported to the ignition site. Third, a firebrand lands on a recipient fuel bed and then ignition occurs if the criteria for ignition is met~\cite{Babrauskas2003}. When considering the possibility of ignition during a fire event all three of processes must be considered as each process significantly influences the properties of an ember landing on a fuel bed. For example, the number, energy content, and size of embers that may land on a fuel bed is significantly influence by the surrounding foliage~\cite{Hudson2020EffectsScale} and moisture content of the foliage~\cite{Adusumilli2021FirebrandShrub} which introduces significant variability in ignition sources a fuel bed may encounter. Similarly, the winds transporting a firebrand significantly influence the path of travel and time of flight which influences the energy content~\cite{Sardoy2007, MatvienkoOVA2016} and localized winds around structures may lead to firebrand accumulation~\cite{Suzuki2020a, Suzuki2017a, Suzuki2015}. Unfortunately, the generation and ignition processes are significantly less understood than the transport processes~\cite{Manzello2020}. The work contained in this dissertation is focused on the ignition process of the fuel bed by a firebrand and does not consider the generation of firebrands. For further information on firebrand generation processes the reader is referred to review papers considering this process by Fernandez-Pello~\cite{Fernandez-Pello2017}, Manzello et al.~\cite{Manzello2020} and Babrauskas~\cite{Babrauskas2003}.

    Firebrands have been identified as a source of structure loss as early as the 1600s~\cite{Suzuki2020} however studies of firebrand ignition have only begun recently. Perhaps one of the first studies published to specifically address firebrands within the scope of structure loss was conducted by Waterman in 1969~\cite{Waterman1969ExperiemntalGeneration}, however research remained sparse until the early 2000s. Early research and a significant portion of current research has been conducted by dropping combusting firebrands~\cite{Ellis2011, Ellis2015, Ganteaume2009, Manzello2006, Manzello2006a} (either flaming or glowing) or hot metal particles~\cite{Wang2017, Urban2017, Fernandez-Pello2015, Hadden2011} onto fuel beds in varying configurations that simulate conditions in a wildfire and observing ignition or no ignition outcomes. These works have identified some parameters that appear to control ignition but models created from these studies only provide qualitative predictions of ignition. The lack of models of quantitative accuracy was suggested by Finney et al. in 2013 who identified the lack of a fundamental understanding of what processes occur and how they interact as a primary gap in the knowledge of ignition~\cite{Finney2013}. A similar need was identified by Manzello et al. in 2018~\cite{Manzello2018} and again in 2020~\cite{Manzello2020}. As fires increase in severity, and the WUI continues to expand the number of homes lost to wildfires is likely to significantly increase in the near future. Thus, the need for accurate predictions of ignition has never been higher.

\section{Modeling Efforts}
    Creating a model with quantitative accuracy of ignition requires a through understanding of the processes and relative importance of each processes leading up to ignition. Figure~\ref{fig:ignitionDiagram} illustrates the different factors that can influence the ignition process. To better understand how these parameters effect the ignition process can be considered as a series of three processes.  The first is the heat transfer to the fuel bed from the ember. The second is the pyrolysis of the fuel bed material and the subsequent release of pyrolyzates into the air above the fuel bed. The third step is the mixing of the pyrolyzates with air above the fuel bed and subsequent gas phase ignition. It is important to note that all three of these processes may occur simultaneously throughout the ignition process and are often interconnected. For example, as heat is transferred from the firebrand to the fuel bed the thermal conductivity of the fuel bed changes with temperature both before and after pyrolysis begins~\cite{Fjellerup2003}. Furthermore, the heat transfer mode and thermal resistance between the firebrand and fuel continuously changes as mass is released due to pyrolysis which may influence ignition~\cite{Yang2016EffectParticle}. The interconnection of the processes that occur during ignition and the inherent variability of fuel beds and biomass fuels in general, the quantification of parameters that influence ignition is quite challenging. Despite these challenges, results of previous research have identified influential parameters that have consistently been observed to control ignition. This work considers these parameters as belonging to one of three categories namely, thermo-physical properties of the fuel bed, energy supplied by the firebrand, and environmental conditions near the ignition point. Detailed reviews of each category are provided in the relevant manuscripts and are not reproduced here. The remaining sections of this review address the predictive models of ignition that have been created from the current state of knowledge. The accuracy and applicability of each model is discussed in turn and the remaining challenges are identified as they have motivated the objectives of this work. 
        \begin{figure}[hpbt]
            \centering
                \includegraphics[width=\textwidth]{Figures/dissertationBed.pdf}
            \caption{Illustration of key features of fuel bed ignition and parameters/characteristics that may influence ignition. The black region near the firebrand represents char formed as the result of pyrolysis.}
            \label{fig:ignitionDiagram}
        \end{figure}

@@

        \begin{equation}
            r_{cr} = \delta_{cr}\sqrt{\frac{k}{\rho A \Delta H}\frac{RT^{2}_{p0}}{E} exp\left(\frac{E}{RT_{p0}}  \right)}
            \label{eqn:hotspot}
        \end{equation}
    In this equation $\delta_{cr}$ is defined as the critical Frank-Kamenetskii hot spot parameter and is determined from calculations~\cite{Goldshleger1973IgnitionDimensions}. $\delta_{cr}$ is dependent on the ratio of volumetric heat capacities of the ignition source and the material, the activation energy of the material, and the temperature of the hot particle. In Equation~\ref{eqn:hotspot} $k, \rho, A, E$, and $\Delta H$ are the thermal conductivity, density, pre-exponetial factor, activation energy, and heat of combustion of the fuel bed material. $R$ is the universal gas constant and $T_{p0}$ is the initial temperature of the heat source.  Ultimately, Equation~\ref{eqn:hotspot} defines the particle energy needed to raise the temperature of a fuel bed region such that enable thermal runaway and subsequent ignition occur. While relatively simple, the hot spot theory has some limitations that prevent quantitative predictions of ignition in a wildfire setting. A discussion of the shortcomings of the hot spot theory are presented by Thomas~\cite{Thomas1965} and more recently presented by Hadden et al.~\cite{Hadden2011} with respect to wildfires. The two most significant shortcomings addressed by the previous works are neglecting heat losses to the surroundings and considering the pyrolysis and ignition phenomena as single step infinitely fast reactions. Hadden et al. attributes these shortcomings as a significant source of error for quantitative ignition predictions. This assertion is reinforced in a review by Fernandez-Pello~\cite{Fernandez-Pello2017} who concluded that in instances where hot spot theory has been applied only qualitatively accurate predictions of ignition have been obtained. A similar model to hot spot theory but with a more comprehensive inclusion of heat loss from the ember was developed by Yin et al.~\cite{Yin2014}. A correlation between fuel properties, energy transferred to the fuel bed, and the time to ignition was proposed and is shown in Equation~\ref{eqn:yinMC}.
        \begin{equation}
            \sqrt{t_{ig}} \approx \frac{q\sqrt{\frac{\rho_{f}k}{c_{p}}}}{\frac{\rho Z \Delta H_c}{t_b}- h_T(T_f - T_0)}
            \label{eqn:yinMC}
        \end{equation}
    Comparing the correlations presented in Equations~\ref{eqn:hotspot} and \ref{eqn:yinMC}, Equation~\ref{eqn:yinMC} includes properties of the fuel bed ($\rho$, $k$, $c_p$, $Z$, $T_0$) and properties the ember or energy source ($\rho_f$, $\Delta H_c/t_b$, $T_f$) as well as the energy required for ignition ($q$) mirroring the Frank-Kamenetskii critical hot spot diameter. The addition of heat losses to the ambient is the most significant difference from the hotspot parameter and addresses a key shortcoming of the hot spot theory. However, the model does not consider the combustion of the fuel bed other than the critical energy for ignition leaving the second key gap for the hot spot theory unaddressed. Perhaps due to the omission of combustion, this model suffers a similar fate as the hot spot theory and follows qualitative trends but is not quantitative. Ultimately, the hot spot theory and similar models do not capture the complexity and variability inherent to ignition~\cite{Manzello2020}.

>>>>

\chapter{Background}
\label{part:literature}
    This chapter provides a discussion of the background and motivation of this work in further depth than the manuscripts presented later in this dissertation. Here, a general overview of ignitions as a threat to structures and fire spread, as well as the current state of knowledge and predictive capabilities, is communicated. Each manuscript includes a relevant and more detailed literature review pertaining to the specific objective of the chapter.

\section{Widlfires and the WUI}
    Wildfires have been an integral part of many ecosystems for millennia and are often beneficial or required for maintaining a healthy ecosystem. As beneficial as they may be to the natural environment, wildfires pose a significant risk to the built environment. As communities around the world have expanded communities outward into environments that support wildfires (e.g., forests and grasslands), the risk of destruction of structures by wildfires has increased~\cite{Hammer2009DemographicManagement}. In the United States, 91\% of the Wildland Urban Interface (WUI) \nomenclature{WUI}{Wildland Urban Interface} growth of 189,000\si{\kilo\meter\squared} (larger than the state of Washington) development between 1990 and 2010 is occupied by homes~\cite{Radeloff2017} and has continued to grow on recent years. The increased development in these areas means that an excess of 40 million homes are located in areas that may be threatened by a wildfire.

    In addition to the increase of homes in the WUI, the intensity and severity of fires have increased. California is an example of the increased fire severity where 13 of the 20 most destructive fires have occurred since 2010, and 7 of the 20 most destructive occurred in 2020 and 2021~\cite{CALFIRE2018}. Increasingly severe fires are not just limited to California. Texas and Tennessee each had one of the top ten most severe fires in the United States between 2005 and 2020. Colorado, Oregon, Washington, Oklahoma, and Florida have also seen highly destructive fires in recent years~\cite{Barrett2020}. The United States is not unique in the presence of wildfires that destroy homes. Australia's Black Summer in 2019-2020 was one of the most severe fire seasons in the country's history~\cite{Levin2021Unveiling2019/2020} where over 3000 homes and greater than 46 million acres burned~\cite{Filkov2020ImpactTrends}, which is approximately equivalent to 75\% of Oregon's land area. The Mediterranean countries of Portugal, Spain, France, Italy, and Greece are also either experiencing greater losses or are a risk for greater losses due to wildfires in recent years. Similar trends are also present in many Asian and South American countries~\cite{Manzello2018}. Wildfires that threaten lives, homes, and structures are truly a global phenomenon, and it is essential to understand how homes are ignited and destroyed in wildfires to better prevent losses from occurring.

\section{The Problem of Ignition}
    Ignition of a home in a WUI fire is attributed to three primary mechanisms; direct flame contact, ignition by radiative heating, and ignition by firebrands. Of these methods, ignition by firebrands is considered to be a primary threat to homes~\cite{Suzuki2021, Mell2010, Manzello2018FORUMResearch}. Case studies of WUI fires have consistently attributed structural losses to firebrand ignition~\cite{Westhaver2017WhyDisaster, Roberts2021} with one study attributing at least 2/3 of home losses to firebrands~\cite{Mell2011}. With the majority of ignitions occurring due to firebrand ignition, it is imperative to understand the mechanism of firebrand ignition to reduce home losses. An analysis of structures burned in California wildfires from 2013 to 2018 found that many of the homes lost would have been considered "fire-safe" by current standards but were still ignited by embers~\cite{Syphard2019}. Codes, standards, and best practices built upon a better understanding of ignition by firebrands would help provide better resilience to firebrand attacks~\cite{Manzello2020} reducing the number of homes lost in wildfires. Examples of these changes may include adjustments in building and landscaping materials favoring the selection of harder to ignite species or housing geometry constraints to reduce the collection of flammable materials or embers. Without a comprehensive knowledge of the ignition process, creating the codes and standards to reduce the risk of home loss is challenging at best and may even be counterproductive.

    The general mechanism for firebrand ignition is considered to be a three-step process that starts within the fire itself, often at a significant distance from the ignition point. First, a firebrand is generated as burning material breaks off or is lofted into the air. Second, the firebrand is then lofted through the air and transported to the ignition site. Third, a firebrand lands on a recipient fuel bed, and then ignition occurs if the criteria for ignition are met~\cite{Babrauskas2003}. When considering the possibility of ignition during a fire event, all three processes must be considered as each process significantly influences the properties of an ember landing on a fuel bed. For example, the number, energy content, and size of embers that may land on a fuel bed are significantly influenced by the surrounding foliage~\cite{Hudson2020EffectsScale} and moisture content of the foliage~\cite{Adusumilli2021FirebrandShrub} which introduces significant variability in ignition sources a fuel bed may encounter. Similarly, the winds transporting a firebrand significantly influence the path of travel and time of flight, which influences the energy content~\cite{Sardoy2007, MatvienkoOVA2016} and localized winds around structures may lead to firebrand accumulation~\cite{Suzuki2020a, Suzuki2017a, Suzuki2015}. Unfortunately, the generation and ignition processes are significantly less understood than the transport processes~\cite{Manzello2020}. The work contained in this dissertation is focused on the ignition process of the fuel bed by a firebrand and does not consider the generation of firebrands. For further information on firebrand generation processes, the reader is referred to review papers considering this process by Fernandez-Pello~\cite{Fernandez-Pello2017}, Manzello et al.~\cite{Manzello2020} and Babrauskas~\cite{Babrauskas2003}.

    Firebrands have been identified as a source of structure loss as early as the 1600s~\cite{Suzuki2020} however, studies of firebrand ignition have only begun recently. Perhaps one of the first studies published to specifically address firebrands within the scope of structure loss was conducted by Waterman in 1969~\cite{Waterman1969ExperiemntalGeneration}. However, research remained sparse until the early 2000s. Early research and a significant portion of current research has been conducted by dropping combusting firebrands~\cite{Ellis2011, Ellis2015, Ganteaume2009, Manzello2006, Manzello2006a} (either flaming or glowing) or hot metal particles~\cite{Wang2017, Urban2017, Fernandez-Pello2015, Hadden2011} onto fuel beds in varying configurations that simulate conditions in a wildfire and observing ignition or no ignition outcomes. These works have identified some parameters that appear to control ignition but models created from these studies only provide qualitative predictions of ignition. The lack of models of quantitative accuracy was suggested by Finney et al. in 2013, who identified the lack of a fundamental understanding of what processes occur and how they interact as a primary gap in the knowledge of ignition~\cite{Finney2013}. A similar need was identified by Manzello et al. in 2018~\cite{Manzello2018} and again in 2020~\cite{Manzello2020}. As fires increase in severity and the WUI continues to expand, the number of homes lost to wildfires is likely to significantly increase in the near future. Thus, the need for accurate predictions of ignition has never been higher.

\section{Modeling Efforts}
    Creating a model with quantitative accuracy of ignition requires a thorough understanding of the processes and relative importance of each process leading up to ignition. Figure~\ref{fig:ignitionDiagram} illustrates the different factors that can influence the ignition process. To better understand how these parameters affect the ignition process can be considered as a series of three processes. The first is the heat transfer to the fuel bed from the ember. The second is the pyrolysis of the fuel bed material and the subsequent release of pyrolyzates into the air above the fuel bed. The third step is the mixing of the pyrolyzates with air above the fuel bed and subsequent gas-phase ignition. It is important to note that all three of these processes may occur simultaneously throughout the ignition process and are often interconnected. For example, as heat is transferred from the firebrand to the fuel bed, the thermal conductivity of the fuel bed changes with temperature both before and after pyrolysis begins~\cite{Fjellerup2003}. Furthermore, the heat transfer mode and thermal resistance between the firebrand and fuel continuously change as mass is released due to pyrolysis, which may influence ignition~\cite{Yang2016EffectParticle}. With the interconnection of the processes that occur during ignition and the inherent variability of fuel beds and biomass fuels in general, the quantification of parameters that influence ignition is quite challenging. Despite these challenges, results of previous research have identified influential parameters that have consistently been observed to control ignition. This work considers these parameters as belonging to one of three categories, namely, thermo-physical properties of the fuel bed, energy supplied by the firebrand, and environmental conditions near the ignition point. Detailed reviews of each category are provided in the relevant manuscripts and are not reproduced here. The remaining sections of this review address the predictive models of ignition that have been created from the current state of knowledge. The accuracy and applicability of each model are discussed in turn, and the remaining challenges are identified as they have motivated the objectives of this work. 
        \begin{figure}[hpbt]
            \centering
                \includegraphics[width=\textwidth]{Figures/dissertationBed.pdf}
            \caption{Illustration of key features of fuel bed ignition and parameters/characteristics that may influence ignition. The black region near the firebrand represents char formed as the result of pyrolysis.}
            \label{fig:ignitionDiagram}
        \end{figure}

@@

        \begin{equation}
            r_{cr} = \delta_{cr}\sqrt{\frac{k}{\rho A \Delta H}\frac{RT^{2}_{p0}}{E} exp\left(\frac{E}{RT_{p0}}  \right)}
            \label{eqn:hotspot}
        \end{equation}
    In this equation, $\delta_{cr}$ is defined as the critical Frank-Kamenetskii hot spot parameter and is determined from calculations~\cite{Goldshleger1973IgnitionDimensions}. $\delta_{cr}$ is dependent on the ratio of volumetric heat capacities of the ignition source and the material, the activation energy of the material, and the temperature of the hot particle. In Equation~\ref{eqn:hotspot} $k, \rho A, E$, and $\Delta H$ are the thermal conductivity, density, pre-exponential factor, activation energy, and heat of combustion of the fuel bed material. $R$ is the universal gas constant, and $T_{p0}$ is the initial temperature of the heat source. Ultimately, Equation~\ref{eqn:hotspot} defines the particle energy needed to raise the temperature of a fuel bed region such that thermal runaway and subsequent ignition occurs. While relatively simple, the hot spot theory has some limitations that prevent quantitative predictions of ignition in a wildfire setting. A discussion of the shortcomings of the hot spot theory is presented by Thomas~\cite{Thomas1965} and more recently presented by Hadden et al.~\cite{Hadden2011} with respect to wildfires. The two most significant shortcomings addressed by the previous works are neglecting heat losses to the surroundings and considering the pyrolysis and ignition phenomena as single-step, infinitely fast reactions. Hadden et al. attribute these shortcomings as a significant source of error for quantitative ignition predictions. This assertion is reinforced in a review by Fernandez-Pello~\cite{Fernandez-Pello2017} who concluded that in instances where hot spot theory has been applied, only qualitatively accurate predictions of ignition have been obtained. A similar model to hot spot theory but with more comprehensive inclusion of heat loss from the ember was developed by Yin et al.~\cite{Yin2014}. A correlation between fuel properties, the energy transferred to the fuel bed, and the time to ignition was proposed and is shown in Equation~\ref{eqn:yinMC}.
        \begin{equation}
            \sqrt{t_{ig}} \approx \frac{q\sqrt{\frac{\rho_{f}k}{c_{p}}}}{\frac{\rho Z \Delta H_c}{t_b}- h_T(T_f - T_0)}
            \label{eqn:yinMC}
        \end{equation}
    Comparing the correlations presented in Equations~\ref{eqn:hotspot} and \ref{eqn:yinMC}, Equation~\ref{eqn:yinMC} includes properties of the fuel bed ($\rho$, $k$, $c_p$, $Z$, $T_0$) and properties the ember or energy source ($\rho_f$, $\Delta H_c/t_b$, $T_f$) as well as the energy required for ignition ($q$) mirroring the Frank-Kamenetskii critical hot spot diameter. The addition of heat losses to the ambient is the most significant difference from the hotspot parameter and addresses a key shortcoming of the hot spot theory. However, the model does not consider the combustion of the fuel bed other than the critical energy for ignition leaving the second key gap for the hot spot theory unaddressed. Perhaps due to the omission of combustion, this model suffers a similar fate as the hot spot theory and follows qualitative trends but is not quantitative. Ultimately, the hot spot theory and similar models do not capture the complexity and variability inherent to ignition~\cite{Manzello2020}.
